Context-dependent consequences of including lagged effects in demographic models

### 2023-07-19 docs/paper.Rmd
removed Renv on startup, fixed render to pdf

      email: embruna@ufl.edu
      institute: [UF_WEC, UF_LAS, BDFFP]
      correspondence: false
institute:
  - UF_WEC: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - UF_LAS: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530  
  - BDFFP: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    # bookdown::pdf_document2:
    #   floatsintext: no
    #   linenumbers: yes
    #   draft: yes
    #   mask: no
    #   figurelist: no
    #   tablelist: no
    #   footnotelist: no
    #   fontsize: 12pt
    #   toc: false
    #   number_sections: no
    #   fig_caption: yes
    bookdown::html_document2: 
      theme: 
        version: 4
        bootswatch: journal
      number_sections: no
      fig_caption: yes
      pandoc_args:
      - --lua-filter=templates/scholarly-metadata.lua
      - --lua-filter=templates/author-info-blocks.lua
      - -Fpandoc-crossref #required for cross-ref equations
    bookdown::word_document2:
      linenumbers: yes
      number_sections: no
      fig_caption: yes
      reference_docx: "templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=templates/scholarly-metadata.lua
      - --lua-filter=templates/author-info-blocks.lua
      - -Fpandoc-crossref #required for cross-ref equations
linkcolor: blue
bibliography: references.bib
csl: "templates/the-american-naturalist.csl" # Insert path for the bib-style
keywords: |
  integral projection models; environmental stochasticity; lagged effects
highlights: |
  These are the highlights

@@

# header-includes:
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyfoot{}
  - \fancyhead[R]{p. \thepage}
  - \fancyhead[L]{Draft - `r format(Sys.time(), '%d %B %Y')`}
  - \usepackage{sectsty} \sectionfont{\centering}
  - \DeclareUnicodeCharacter{U+2206}{$\delta$}
  # - \usepackage{sectsty} \subsectionfont{\centering}
  # - \AtBeginDocument{\let\maketitle\relax}
 #  - \usepackage{float} #to keep figure in the location placed instead of end
 #  - \floatplacement{figure}{H}
 #  - \usepackage{booktabs}
 #  - \usepackage{setspace} # line spacing
 #  - \raggedbottom
 # # - \usepackage{endfloat} #[nomarkers] excludes the {insert figure x around here] from main text. The others exclude the list of tables and figures. https://cs.brown.edu/about/system/managed/latex/doc/endfloat.pdf
 #  - \usepackage{tabu}
 #  - \usepackage{makecell}
 #  - \usepackage{pdflscape}
 #  - \usepackage{longtable}
 #  - \newcommand{\blandscape}{\begin{landscape}}
 #  - \newcommand{\elandscape}{\end{landscape}}
 #  - \setlength\parindent{22pt}
---

@@

1.  Demographic models are widely used for all kinds of stuff

2.  It has long been recognized that there is the potential for lagged effects.
    Now there's evidence for that [@eversLaggedDormantSeason2021; @scottDelayedEffectsClimate2022].

@@

# Tables

@@

  arrange(habitat, model) %>% 
  select(
    "Habitat" = "habitat",
    "Vital Rate" = "model",
    "Effect of Environment" = "env_effect_incl",
    #"Model" = "RHS",
    "edf" = "df",
    "∆AIC" = "dAIC"
  ) %>%

pandoc.table(
  justify = c( "left", "left","left", "right",  "right"),
  full_width = FALSE,
  caption = "(\\#tab:aic) Comparison of vital rate models used to build IPM.  The 'Effect of Environment' column describes how environmental effects were included in models. Those with 'none' were used to build deterministic IPMs; those with a random effect of year were used to build stochastic, kernel-resampled IPMs; and those with a distributed lag non-linear model (DLNM) were used to build stochastic, parameter-resampled IPMs. 'edf' is the estimated degrees of freedom of the penalized GAM. ∆AIC is calculated within each habitat and vital rate combination.  ∆AIC within 2 indicates models are equivalent."
)
```

@@

# Figures

```{r lifecycle}
#| fig.cap = "Lifecycle diagram of *Heliconia acuminata*. Each transition is associated with an equation for a vital rate function.  The functions shown on the diagram correspond to those used to construct a general, density-independent, deterministic IPM. The table below shows the equivalent equations for stochastic, kernel-resampled IPMs and stochastic, parameter-resampled IPMs."

@@

```{r pop-states}
#| fig.cap = "Relative proportions of plant sizes in the first 250 iterations of the IPM simulations. Stacked area charts (A) show the relative size/stage distribution of plants in continuous forest (CF, top row) and forest fragments (FF, bottom row) in each of the three IPMs (columns). The proportion of each size class in CF and FF for each iteration is shown in B with the first 30 iterations removed to not include transient dynamics.  A 1:1 line is plotted in black.  Size categories include seedlings (a discrete category in the IPMs), pre-reproductive 1 (log(size) 0–2.5) that have low average survival (< 0.9) and a near 0 probability of flowering, pre-reproductive 2 (log(size) 2.5–4.5) that have a higher average survival probabilty (> 0.8) and a near 0 probability of flowering, reproductive 1 (log(size) 4.5–6) that have a high average survival probability (>0.95) and a lower flowering probability (< 0.25), and reproductive 2 (log(size) 6+) that have a high average survival probability (>0.95) and higher flowering proability (> 0.2)."

@@

>>>>

      email: embruna@ufl.edu
      institute: [UF_WEC, UF_LAS, BDFFP]
      correspondence: false
institute:
  - UF_WEC: Department of Wildlife Ecology and Conservation, University of Florida, Gainesville, Florida 32611-0430
  - UF_LAS: Center for Latin American Studies, University of Florida, Gainesville, Florida 32611-5530  
  - BDFFP: Biological Dynamics of Forest Fragments Project, INPA-PDBFF, CP 478, Manaus, Amazonas 69011-970 Brazil
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    # bookdown::html_document2:
    #   theme:
    #     version: 4
    #     bootswatch: journal
    #   number_sections: no
    #   fig_caption: yes
    #   pandoc_args:
    #   - --lua-filter=templates/scholarly-metadata.lua
    #   - --lua-filter=templates/author-info-blocks.lua
      # - -Fpandoc-crossref #required for cross-ref equations
    bookdown::pdf_document2:
      floatsintext: no
      linenumbers: yes
      draft: yes
      mask: no
      figurelist: no
      tablelist: no
      footnotelist: no
      fontsize: 12pt
      toc: false
      number_sections: no
      fig_caption: yes
      pandoc_args:
      - --lua-filter=templates/scholarly-metadata.lua
      - --lua-filter=templates/author-info-blocks.lua
    # bookdown::word_document2:
    #   linenumbers: yes
    #   number_sections: no
    #   fig_caption: yes
    #   reference_docx: "templates/template.docx" # Insert path for the DOCX file
    #   pandoc_args:
    #   - --lua-filter=templates/scholarly-metadata.lua
    #   - --lua-filter=templates/author-info-blocks.lua
    #   - -Fpandoc-crossref #required for cross-ref equations
linkcolor: blue
bibliography: references.bib
csl: "templates/the-american-naturalist.csl" # Insert path for the bib-style
keywords: |
  integral projection models; environmental stochasticity; lagged effects
highlights: |
  These are the highlights

@@

# header-includes:
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyfoot{}
  - \fancyhead[R]{p. \thepage}
  - \fancyhead[L]{Draft - `r format(Sys.time(), '%d %B %Y')`}
  - \usepackage{sectsty} \sectionfont{\centering}
  - \usepackage[utf8]{inputenc} # This loads package to convert unicode 
  - \DeclareUnicodeCharacter{394}{$\Delta$} # convert unicode \u0394 to Delta
  # - \usepackage{sectsty} \subsectionfont{\centering}
  # - \AtBeginDocument{\let\maketitle\relax}
 #  - \usepackage{float} #to keep figure in the location placed instead of end
 #  - \floatplacement{figure}{H}
 #  - \usepackage{booktabs}
 #  - \usepackage{setspace} # line spacing
 #  - \raggedbottom
 # # - \usepackage{endfloat} #[nomarkers] excludes the {insert figure x around here] from main text. The others exclude the list of tables and figures. https://cs.brown.edu/about/system/managed/latex/doc/endfloat.pdf
 #  - \usepackage{tabu}
 #  - \usepackage{makecell}
 #  - \usepackage{pdflscape}
 #  - \usepackage{longtable}
 #  - \newcommand{\blandscape}{\begin{landscape}}
 #  - \newcommand{\elandscape}{\end{landscape}}
 #  - \setlength\parindent{22pt}
---

@@

1.  Demographic models, such as IPMs and matrix models, are widely used for all kinds of stuff.

2.  It has long been recognized that there is the potential for lagged effects.
    Including lagged effects in models has been a major technical challenge, but there are now multiple approaches for doing so.
    The studies assessing the potential for lagged effects on vital rates find that they indeed appear to be prevalent [@eversLaggedDormantSeason2021; @scottDelayedEffectsClimate2022].

@@

<!-- # Tables -->

@@

  arrange(habitat, model) %>% 
  select(
    "Habitat" = "habitat",
    "Vital Rate" = "model",
    "Effect of Environment" = "env_effect_incl",
    #"Model" = "RHS",
    "edf" = "df",
    "\u0394AIC" = "dAIC"
  ) %>%

pandoc.table(
  justify = c( "left", "left","left", "right",  "right"),
  full_width = FALSE,
  caption = "(\\#tab:aic) Comparison of vital rate models used to build IPM.  The 'Effect of Environment' column describes how environmental effects were included in models. Those with 'none' were used to build deterministic IPMs; those with a random effect of year were used to build stochastic, kernel-resampled IPMs; and those with a distributed lag non-linear model (DLNM) were used to build stochastic, parameter-resampled IPMs. 'edf' is the estimated degrees of freedom of the penalized GAM. \u0394AIC is calculated within each habitat and vital rate combination.  \u0394AIC within 2 indicates models are equivalent."
)
```

\newpage

@@

\newpage

<!-- # Figures -->

```{r lifecycle,  echo=FALSE, out.width="85%",fig.align='center'}
#| fig.cap = "Lifecycle diagram of *Heliconia acuminata*. Each transition is associated with an equation for a vital rate function.  The functions shown on the diagram correspond to those used to construct a general, density-independent, deterministic IPM. The table below shows the equivalent equations for stochastic, kernel-resampled IPMs and stochastic, parameter-resampled IPMs."

@@

\newpage

```{r pop-states,  echo=FALSE, out.width="85%",fig.align='center'}
#| fig.cap = "Relative proportions of plant sizes in the first 250 iterations of the IPM simulations. Stacked area charts (A) show the relative size/stage distribution of plants in continuous forest (CF, top row) and forest fragments (FF, bottom row) in each of the three IPMs (columns). The proportion of each size class in CF and FF for each iteration is shown in B with the first 30 iterations removed to not include transient dynamics.  A 1:1 line is plotted in black.  Size categories include seedlings (a discrete category in the IPMs), pre-reproductive 1 (log(size) 0–2.5) that have low average survival (< 0.9) and a near 0 probability of flowering, pre-reproductive 2 (log(size) 2.5–4.5) that have a higher average survival probabilty (> 0.8) and a near 0 probability of flowering, reproductive 1 (log(size) 4.5–6) that have a high average survival probability (>0.95) and a lower flowering probability (< 0.25), and reproductive 2 (log(size) 6+) that have a high average survival probability (>0.95) and higher flowering proability (> 0.2)."

@@

\newpage
